A Critical Appraisal of Zero-Multiple Method in Classical Water Simulation

### 2014-11-03 writedown/zm/paper.water/main.tex
further polish the paper.

A very short overview of the existing electrostatic interaction methods.
Long-range treatment, Ewald and derived methods like SPME.
Short-range treatment.

@@

A short mention of the group cut-off v.s.~the atom cut-off method. (will discuss the implementation in gromacs later)

@@

In order to investigate the performance of ZM method, we simulation
the water system by using both SPME and ZM method, and calculate a set
of \emph{benchmark quantities}.  This set covers a broad range of
water properties that are of general interest, and includes the
structure property (radial distribution function), the thermodynamic
properties (pressure, excess chemical potential, constant
volume/pressure heat capacity, isothermal compressibility, thermal
expansion coefficient), dielectric properties (dielectric constant and
Kirkwood-G factor) and dynamical properties (diffusion coefficient and
viscosity).  By comparing ZM to SPME, a good consistency in computing
the benchmark quantities would indicate a good accuracy of the ZM
method.  Moreover, being aware of the weakness of the ZM method also
prevents the application the method in wrong applications.  The ZM
method is also compared to the traditional short-range method: RF.
The definitions for computing these quantities are provided in details
in Appendix.~\ref{appendix:benchmark}.

@@

In this work, we investigate the water system modeled by
TIP3P~\cite{jorgensen1983comparison} point-charge water model.
All simulations were performed by Gromacs~4.6~\cite{hess2008gromacs, pronk2013gromacs} complied by GNU C compiler with single float precision.
The size of the systems and corresponding number of molecules used for MD simulations are summarized in Tab.~\ref{tab:tmp1}.
The benchmark quantities were calculated in either  canonical (NVT) or isothermal-isobaric (NPT) ensemble, with equilibrium temperature $T=300$~K.
In order to simulate at the desired temperature,
the systems $\systemsb$ and $\systemsbp$ were coupled to either Nose-Hoover~\cite{nose1984molecular,hoover1985canonical} thermostat with time-scale $\tau_T = 1.0$~ps,
while systems $\systemmb$, $\systemmbp$, $\systemla$, $\systemlb$ and $\systemlbp$ were coupled to Langevin thermostat with time-scale $\tau_T = 0.1$~ps.
In the NPT simulations, the systems were further coupled
to the Parrinello-Rahman barostat~\cite{parrinello1980crystal,parrinello1981polymorphic} (in Gromacs implementation)
with time-scale $\tau_P = 0.5$~ps to equilibriate the system at 1~Bar.
All systems were integrated with leap-frog scheme at the time-step of 0.001~ps.
For the ZM and SPME methods,
the neighbor lists were updated every 5 time-steps.
The ``\texttt{group}'' cut-off scheme in Gromacs was used. In order to mimic the atomistic
cut-off manner, the radius for neighbor list building was set 0.3~nm
larger than the cut-off radius (the maximum of Coulomb cut-off and the van der Waals cut-off).
To make sure the van der Waals interactions were cut-off in a smooth way,
they are modified in a shell of $0.05$~nm width near the cut-off sphere by the Gromacs ``\texttt{shift}'' method.
The ZM method was implemented in Gromacs with the tabulated electrostatic interaction (Home-made codes for generating the interaction tables).
The short-range part of SPME method is also smoothed in a shell of $0.05$~nm width  by the Gromacs ``\texttt{pme-switch}'' method.
The reciprocal space
grid spacing is set to 0.06~nm, with B-Spline interpolation order of 6.
These parameters are chosen to be more accurate than the Gromacs default choices (grid spacing 0.12~nm and interpolation order 4).
The splitting parameter is optimized by the error estimate for the SPME method, which is implemented in Gromacs by tool ``\texttt{g\_pme\_error}''~\cite{wang2010optimizing}.
The optimized real space and reciprocal space force computing errors were equally distributed as $1.4\times10^{-4}$~kJ/(mol nm).
In the simulations of the RF method, the interactions were cut-off in the group manner,
so the radius for neighbor list building is set identical to the cut-off radius.
No cut-off smoothing was applied in the cases of RF simulation.

\begin{table}
  \centering
  \caption{A list of systems simulated in this work. We report, from left to right, the number of water
    molecules in the system (TIP3P model), equilibrium box size, length of equilibriation, length of productive trajectory, the thermostats and the corresponding coupling time-scales.
  }
  \vskip .5cm
  \begin{tabular*}{0.9\textwidth}{@{\extracolsep{\fill}}c rrrrrr}\hline\hline
    System      & $N$ & $V$ [$\textrm{nm}^3$]  & $T_{eq}$ [ps] & Traj. Length [ps] & Thermostat & $\tau$ [ps]  \\\hline
    $\systemsbp$         &   2,500        & $4.24^3$              & --           &  6,000        & Nose-Hoover & 1.0\\
    $\systemsb$          & 2,500          & $4.24^3$              & 3,000        & 27,000        & Nose-Hoover & 1.0 \\\hline
    $\systemmbp$         &   4,500        & $5.15^3$              & --           &  6,000        & Langevin & 0.1\\
    $\systemmb$          & 4,500          & $5.15^3$              & 3,000        & 17,000        & Langevin & 0.1\\\hline
    $\systemla$          &   13,824       & $7.49^3$              & 200          & 1,800         & Langevin & 0.1\\
    $\systemlbp$         &   13,824       & $7.49^3$              & --           & 10,000        & Langevin & 0.1\\
    $\systemlb$          &   13,824       & $7.49^3$              & 3,000        & 17,000        & Langevin & 0.1\\
    \hline\hline
  \end{tabular*}
  \label{tab:tmp1}
\end{table}

The initial configurations were prepared by in
the following steps:
(1) An NPT simulation of system $\systemla$ with SPME
electrostatic was performed. The equilibrium density (984.1~$\textrm{kg}/\textrm{m}^{-3}$) of the system is calculated
from this simulation.
(2) Initial configurations that contains 2500, 4500 and
13824 molecules were generated at the calculated equilibrium
density.
(3) NVT simulations with SPME electrostatic for system $\systemsbp$, $\systemmbp$
and $\systemlbp$ are performed.
(4) The output configurations of $\systemsbp$ and $\systemmbp$ were used as
initial configurations for system $\systemsb$ and $\systemmb$, respectively.  The output configuration of
$\systemlbp$ was used as initial configurations for system $\systemla$ and $\systemlb$.
% The SPME
% parameters for all simulations reported by this work is: real space
% cut-off 1.90~nm.

@@

The smallest system $\systemsb$ was adopted,
because for the quantities that are computed  by estimating fluctuations (in this
work heat capacities, compressibility, thermal expansion coefficient, dielectric constant, diffusion constant and
viscosity are fluctuations. See details in Appendix~\ref{appendix:benchmark}), simply increase the size of the
system does not improve the accuracy of the 
simulation~\cite{milchev1986fluctuations,ferrenberg1991statistical}.
The optimal way of measuring these properties
is to simulate with smaller systems (as far as the finite size effect is negligible) and run longer simulations for longer time-averages.
The reason of using Nose-Hoover thermostat is that
it perturbs the dynamics of the system only slightly, while the
Langevin thermostat is likely to substantially change the dynamics
when the coupling is very strong, therefore, the former was used in computing the dynamical
properties like the diffusion constant and viscosity.
The quantities that
are ensemble averages (in this work, pressure, chemical potential, radial distribution function and Kirkwood-G factor are of this type)
were computed by the largest system $\systemlb$,
in order to achieve higher parallel efficiency.
The system $\systemmb$ is simulated to investigate the size-dependency of the
Kirkwood-G factor~\cite{vanderSpoel2006origin}.

@@

The equilibrium density of the system was determined by an NPT SPME
simulation.  In the NVT simulation the SPME electrostatic, computed
pressure ($-0.4\pm 1.5$~Bar) is reproducing the target pressure of
1.0~Bar.  The accuracy of ZM method, can be investigate by checking if
the NVT pressure is consistent with the target pressure.  In
Fig.~\ref{fig:pres-comp}, the pressure calculated by ZM method of
order $l=1,2,3$ and 4 are plotted against the cut-off radius.
All the  simulations are done with system $\systemla$.
It is concluded from the Figure that all ZM methods converges to the
target pressure at large enough cut-off radius.
In general, the higher order of ZM presents an larger error in the pressure calculation than
the lower order ZM method.
Higher order are also more sensitive in the  cut-off radius dependency.
ZM $l=1$ tends to under-estimate the pressure, while ZM $l=2,3$ and 4 tends to over-estimate.
The ZM method is also compared with
the RF method using the experimental dielectric constant $\varepsilon_{\textrm{rf}} = 80$, 
from which it is concluded that the accuracy of ZM in pressure calculation is much higher
than the RF method.
It should be noted that even with the largest system of 13824 water molecules,
the fluctuation (measured by standard deviation) of the pressure is roughly 100~Bar, which means that it is
very difficult achieve very accurate estimate for this quantity. In practice a pressure error of 20~Bar is
still acceptable.

@@

The effect of using different splitting parameter on the convergence
of the pressure is investigate in Fig.~\ref{fig:pres-l1}.
In general, the higher order indicates less significant dependency on the splitting parameter.
In the case of $l=1$, the splitting parameter 0.0 is consistent with  $0.5\,\textrm{nm}^{-1}$, and
by using higher splitting parameter, the pressure shifts to the higher pressure side.
For $l=2,3$ and 4, the convergence is not sensitive to $\alpha$ up to  $1.0\,\textrm{nm}^{-1}$.
More importantly, for all orders,
the pressure does not converges to the desired value 1~Bar by using $\alpha = 1.5$ or $2.0\,\textrm{nm}^{-1}$.

@@

The artificial oscillation can be substantially improved by using larger splitting parameter $\alpha$.
In Fig.~\ref{fig:rdf-damp} presents the RDF of ZM $l=1$ with splitting parameter from 0.0 to 2.0~$\textrm{nm}^{-1}$.
A clear trend of better RDF by using larger splitting parameter is observed.
At $\alpha = 2.0\ \textrm{nm}^{-1}$ the RDF is almost overlapping with the SPME result.

@@

As a measurement of the dipole correlation as a function of molecular distance, the Kirkwood-G factor
is an important quantity of the water system.
It is well know that the G-factor is system size
dependent~\cite{vanderSpoel2006origin}, therefore we show the G-factor for
both  system $\systemlb$ in Fig.~\ref{fig:gkr}~(a) and a smaller system $\systemmb$
Fig.~\ref{fig:gkr} (b).  In the figures, the statistical uncertainty of
the SPME method is denoted by the red error bars. The statistical uncertainties of the
other methods are essentially of the same size as the SPME, so they are not presented in the figure for
clarity.  In the figure, ZM adopts a cut-off radius of 1.2~nm and a splitting parameter $\alpha = 0.00\,\textrm{nm}^{-1}$.
RF uses a cut-off 1.2~nm and a dielectric constant of 80.

@@

The cut-off and splitting parameter dependency in calculating the Kirkwood-G factor
is presented in Fig.~\ref{fig:gkr-conv}  for ZM method of order $l=3$. With out surprising,
the Kirkwood-G factor converges to the SPME result when using increasing cut-off radius. 
However, unlike the case of calculating RDF, the accuracy Kirkwood-G factor is not improved
with increasing splitting parameter. This fact should be noticed in choosing parameters for
high accuracy calculation of Kirkwood-G factor.

>>>>

A very short overview of the existing electrostatic interaction methods. Two classes in general:
Long-range treatment, e.g.~Ewald and derived methods like SPME.
Short-range treatment, like RF and ZM.

@@

\recheck{If we want to discuss the issue of cut-off method, a short
  introduction of the group cut-off v.s.~the atom cut-off
  method. (will discuss the implementation in gromacs later)}

@@

In order to investigate the performance of ZM method, we simulation
the water system by using both SPME and ZM methods, and calculate a set
of \emph{benchmark quantities}.  This set covers a broad range of
water properties that are of general interest. It includes the
structure property (radial distribution function), the thermodynamic
properties (pressure, excess chemical potential, constant
volume/pressure heat capacity, isothermal compressibility, thermal
expansion coefficient), dielectric properties (dielectric constant and
Kirkwood-G factor) and dynamical properties (diffusion constant and
viscosity).
The definitions and computing methods of these quantities are provided in details
in Appendix.~\ref{appendix:benchmark}.
By comparing ZM to SPME, a good consistency in computing
the benchmark quantities would indicate a good accuracy of the ZM
method.
More importantly, the benchmark tests provide knowledge in the
stituations when the ZM method does not work so well, and help in
finding the way of tuning the working parameters to achieve higher accuracy.
The ZM method is also compared to the traditional short-range method: RF.

@@

In this work, we focus on the water system modeled by
TIP3P~\cite{jorgensen1983comparison} point-charge water model.
% Other point-charge water models differ in the modeling parameters, and
The accuracy of ZM method in other water models is out of the scope of the current work.
However, since the point-charge water models share similar idea, and  only differ
in the model parameters, it is highly probable that ZM works well with other water models if it works
well with TIP3P.

All simulations were performed by Gromacs~4.6.5~\cite{hess2008gromacs, pronk2013gromacs} complied by GNU C compiler with single-precision floating point.
The size of the systems and corresponding number of molecules used for MD simulations are summarized in Tab.~\ref{tab:tmp1}.
The benchmark quantities were calculated in either  canonical (NVT) or isothermal-isobaric (NPT) ensemble, with equilibrium temperature $T=300$~K.
In order to simulate at the desired temperature,
the systems $\systemsb$ and $\systemsbp$ were coupled to Nose-Hoover~\cite{nose1984molecular,hoover1985canonical} thermostat with time-scale $\tau_T = 1.0$~ps,
while systems $\systemmb$, $\systemmbp$, $\systemla$, $\systemlb$ and $\systemlbp$ were coupled to Langevin thermostat with time-scale $\tau_T = 0.1$~ps.
In the NPT simulations, the systems were further coupled
to the Parrinello-Rahman barostat~\cite{parrinello1980crystal,parrinello1981polymorphic} (in Gromacs implementation)
with time-scale $\tau_P = 0.5$~ps to equilibriate the system at 1~Bar.
All systems were integrated with leap-frog scheme at a time-step of 0.001~ps.
The neighbor lists were updated every 5 time-steps.
The ``\texttt{group}'' cut-off scheme in Gromacs was used. In order to mimic the atomistic
cut-off manner, the radius for neighbor list building was set 0.3~nm
larger than the cut-off radius (the maximum of Coulomb cut-off and the van der Waals cut-off).
To make sure the van der Waals interactions were cut-off in a smooth way,
they are modified in a shell of $0.05$~nm width near the cut-off sphere by the Gromacs ``\texttt{shift}'' method.
The ZM method was implemented in Gromacs with the tabulated electrostatic interaction. A home-made code was used to generate the interaction tables.
The short-range part of SPME method is smoothed in a shell of $0.05$~nm width  by the Gromacs ``\texttt{pme-switch}'' method.
The reciprocal space
grid spacing is set to 0.06~nm, with B-Spline interpolation order of 6.
These parameters are chosen to be more accurate than the Gromacs default choices (grid spacing 0.12~nm and interpolation order 4).
The splitting parameter is optimized by the error estimate for the SPME method, which is implemented in Gromacs by tool ``\texttt{g\_pme\_error}''~\cite{wang2010optimizing}.
The optimized real space and reciprocal space force computing errors were equally distributed as $1.4\times10^{-4}$~kJ/(mol nm).
Since the magnitude of the forces are of order $10^{2}$~kJ/(mol nm), this force computing precision almost reaches the limit of the single-precision representation of the floating point numbers.
In the simulations of the RF method, the interactions were cut-off in the group manner,
so the radius for neighbor list building is set identical to the cut-off radius.
No cut-off smoothing of the electrostatic interaction was applied in the cases of RF simulation.

\begin{table}
  \centering
  \caption{A list of systems simulated in this work. We report, from left to right, the number of water
    molecules in the system (TIP3P model), equilibrium box size, length of equilibriation, length of productive trajectory.
  }
  \vskip .5cm
  \begin{tabular*}{0.5\textwidth}{@{\extracolsep{\fill}}c rrrr}\hline\hline
    System      & $N$ & $V$ [$\textrm{nm}^3$]  & $T_{\textrm{eq}}$ [ps] & $T_{\textrm{prod}}$ [ps]  \\\hline
    $\systemsbp$         &   2,500        & $4.24^3$              & --           &  6,000        \\
    $\systemsb$          & 2,500          & $4.24^3$              & 3,000        & 27,000        \\\hline
    $\systemmbp$         &   4,500        & $5.15^3$              & --           &  6,000        \\
    $\systemmb$          & 4,500          & $5.15^3$              & 3,000        & 17,000        \\\hline
    $\systemla$          &   13,824       & $7.49^3$              & 200          & 1,800         \\
    $\systemlbp$         &   13,824       & $7.49^3$              & --           & 10,000        \\
    $\systemlb$          &   13,824       & $7.49^3$              & 3,000        & 17,000        \\
    \hline\hline
  \end{tabular*}
  \label{tab:tmp1}
\end{table}

The initial configurations were prepared by in
the following steps:
(1) An NPT simulation of system $\systemla$ with SPME
electrostatic was performed. The equilibrium density (984.1~$\textrm{kg}/\textrm{m}^{3}$) of the system is calculated
from this simulation.
(2) Initial configurations that contains 2500, 4500 and
13824 molecules were generated at the calculated equilibrium
density.
(3) NVT simulations with SPME electrostatic for system $\systemsbp$, $\systemmbp$
and $\systemlbp$ are performed.
(4) The output configurations of $\systemsbp$ and $\systemmbp$ were used as
initial configurations for system $\systemsb$ and $\systemmb$, respectively.  The output configuration of
$\systemlbp$ was used as initial configurations for system $\systemla$ and $\systemlb$.
% The SPME
% parameters for all simulations reported by this work is: real space
% cut-off 1.90~nm.

@@

The smallest system $\systemsb$ was adopted,
because for the quantities that are computed  by estimating fluctuations (in this
work heat capacities, compressibility, thermal expansion coefficient, dielectric constant, diffusion constant and
viscosity are fluctuations. See details in Appendix~\ref{appendix:benchmark}), simply increase the size of the
system does not improve the accuracy of the 
simulation~\cite{milchev1986fluctuations,ferrenberg1991statistical}.
The optimal way of measuring these properties
is to simulate with smaller systems
% (as far as the finite size effect is negligible)
and run longer simulations for longer time-averages.
The reason of using Nose-Hoover thermostat is that
it perturbs the dynamics of the system only slightly, while the
Langevin thermostat is likely to substantially change the dynamics
when the coupling is very strong, therefore, the former was used in computing the dynamical
properties like the diffusion constant and viscosity.
The quantities that
are ensemble averages (in this work, pressure, chemical potential, radial distribution function and Kirkwood-G factor are of this type)
were computed by the largest system $\systemlb$,
in order to achieve higher parallel efficiency.
The system $\systemmb$ is simulated to investigate the size-dependency of the
Kirkwood-G factor~\cite{vanderSpoel2006origin}.

@@

The equilibrium density of the system was determined by an NPT SPME
simulation.  Then using this density in the NVT simulation with SPME electrostatic, the
pressure ($-0.4\pm 1.5$~Bar) is reproducing the target pressure of
1.0~Bar, which means the system is in desired thermodynamic state.  The accuracy of ZM method, can be investigate by checking if
the NVT pressure is consistent with the target pressure, when the SPME equilibrium density is used.  In
Fig.~\ref{fig:pres-comp}, the pressure calculated by ZM method of
order $l=1,2,3$ and 4 are plotted against the cut-off radius.
All the  simulations are done with system $\systemla$.
It is concluded from the Figure that all ZM methods converges to the
target pressure at large enough cut-off radius.
In general, the higher order of ZM presents an larger error in the pressure calculation than
the lower order ZM method.
Higher order are also more sensitive in the  cut-off radius dependency.
ZM $l=1$ tends to under-estimate the pressure, while ZM $l=2,3$ and 4 tends to over-estimate.
The ZM method is also compared with
the RF method using the experimental dielectric constant $\varepsilon_{\textrm{rf}} = 80$, 
from which it is concluded that the accuracy of ZM in pressure calculation is much higher
than the RF method.
It should be noted that even with the largest system of 13824 water molecules,
the fluctuation (measured by standard deviation) of the pressure is roughly 100~Bar, which means that it is
very difficult achieve very accurate estimate for this quantity. In practice a pressure error of 20~Bar is
still acceptable.

@@

The effect of using different splitting parameter on the convergence
of the pressure is investigate in Fig.~\ref{fig:pres-l1}.
In general, the higher order indicates less significant dependency on the splitting parameter.
In the case of $l=1$, the splitting parameter 0.0 is consistent with  $0.5\,\textrm{nm}^{-1}$, and
by using higher splitting parameter, the pressure shifts to the higher pressure side.
For $l=2,3$ and 4, the convergence is not sensitive to $\alpha$ up to  $1.0\,\textrm{nm}^{-1}$.
More importantly,  by using $\alpha = 1.5$ or $2.0\,\textrm{nm}^{-1}$,
the pressure does not converges to the desired value 1~Bar. Therefore, the high splitting parameter
is not suggested in the simulations where the high pressure accuracy is important.

@@

The artificial oscillation can be substantially improved by using larger splitting parameter $\alpha$.
Fig.~\ref{fig:rdf-damp} presents the RDF of ZM $l=1$ with splitting parameter from 0.0 to 2.0~$\textrm{nm}^{-1}$.
A clear trend of better RDF by using larger splitting parameter is observed.
At $\alpha = 2.0\ \textrm{nm}^{-1}$ the RDF is almost overlapping with the SPME result.

@@

As a measurement of the dipole correlation as a function of molecular distance, the Kirkwood-G factor
is an important quantity of the water system.
It is well know that the G-factor is system size
dependent~\cite{vanderSpoel2006origin}, therefore we show the G-factor for
system $\systemlb$ in Fig.~\ref{fig:gkr}~(a) and a smaller system $\systemmb$
in Fig.~\ref{fig:gkr} (b).  In the figures, the statistical uncertainty of
the SPME method is denoted by the red error bars. The statistical uncertainties of the
other methods are essentially of the same size as the SPME, so they are not presented in the figure for
clarity.  In the figure, ZM adopts a cut-off radius of 1.2~nm and a splitting parameter $\alpha = 0.00\,\textrm{nm}^{-1}$.
RF uses a cut-off 1.2~nm and a dielectric constant of 80.

@@

The cut-off and splitting parameter dependency in calculating the Kirkwood-G factor
is presented in Fig.~\ref{fig:gkr-conv}  for ZM method of order $l=3$. With out surprising,
the Kirkwood-G factor converges to the SPME result when using increasingly large cut-off radius. 
However, unlike the case of calculating RDF, the accuracy Kirkwood-G factor is not improved
with increasing splitting parameter. This fact should be noticed in choosing parameters for
high accuracy calculation of Kirkwood-G factor.
